### 2017-03-30 Manuscripts/cog-psych/cogpsych.tex
minor edits to manuscript

\jlanote{The issue isn't prior knowledge actually -- Bayesian models can handle that gracefully -- it's the rich format of participant responses being 2-d sketches}

\cite{jern2013probabilistic} recently showed that creative generation could be studied in a more controlled manner through the well-developed methods of an artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In Experiments 3 and 4 of their article, participants were exposed to members of experimenter-defined categories of "crystals" varying in size, hue, and saturation. Following a training phase during which the experimenter-defined categories were learned, participants were asked to generate novel categories of crystals. In a finding mirroring that of the \cite{ward1994structured} studies, \citeauthor{jern2013probabilistic} found that participants generated categories with the same distributional properties as the experimenter-defined categories: for example, after training on categories with a positive correlation between the size and saturation features (larger sized crystals were more saturated), participants generated novel categories with the same positive correlation. This finding is notable, as it demonstrates that category generation can be studied in a well-known and highly controlled experimental paradigm. Note that generated exemplars for the novel category automatically contrasted from previous categories because the participants selected the size and saturation, but the exemplar's hue was set (and fixed) to a novel value. As such, hue was ignored in their analyses and may have satisfied any constraints contrast may have on category generation.

\jlanote{It is hierarchical Bayesian, not Bayesian hierarchical.}

@@

% Here's what it used to be:
% Their model represents categories as multivariate normal distributions within the space, and from exposure to category members the model induces a domain-wide representation of the distributional commonalities among the observed categories. When generating a new category, the model samples a new distribution from the domain, and thus the distributional properties of observed categories will be preserved within the generates category.

% \jlanote{Yeah, Here's how I would this section should be organized. Top is general behavioral findings/phenomena. Then subsections for each theory, which first is described in broad theoretical terms and then how that is formalized into a model}

\cite{jern2013probabilistic} additionally tested a "copy-and-tweak" model that broadly resembles the earlier "Path of Least Resistance" account. The core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}: The model generates new items according to their similarity to known members of the target category. This model provided a poor account of the observed data. Indeed, the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, its application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims little about how generated categories should contrast with what is already known. In their simulations, the model was only tested on generation after the learner had been exposed to members of the target category, and so the model's ability to generate a new category was not evaluated. However, the model's generation is based exclusively on similarity to known members of the target category; when there are no members of the target category, generation is presumably random.

@@

\nbcnote{Maybe note something about how PACKER's basis in the exemplar view means that we're explaining category generation using well known ideas, and that exemplar models are viewed as highly principled.}

\jlanote{Yeah, I think this would make a nice subsection in the general discussion.}

Both PACKER and the GCM simulate categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and model performance is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{attneave1950,shepard1957stimulus,shepard1987toward}:
\jlanote{attneave was first (though he was inspired by hypotheses made by householder and others, so it may date back even earlier). he framed it in terms of sum of log distances, but it's equivalent. funnily he said the biggest issue is violations of the triangle inequality via examples attributed to James. i can send you the article if you want....}
\begin{equation}
s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k \left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, attention will be distributed uniformly in our simulations (unless otherwise noted). The value of $r$ depends on the nature of the experimental conditions being simulated: $r=1$ is appropriate for separable dimensions, whereas $r=2$ is appropriate for integral dimensions \citep[see][]{shepard1964attention,garner1974processing}. In our simulations, we set $r=1$ due to the separable nature of the stimulus dimensions used in our experiments (see Figure \ref{fig:e1-conditions}).

PACKER (as well as its name) was in part inspired by earlier work from the categorization literature \citep{hidaka2011packing}. They argued that natural categories "pack" the values of features such that they fill the space with as much distance apart from one another, while maintaining items within the same category close together. Inspired by this idea, PACKER proposes that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories: the most desirable generation candidates are similar to members of the target category and not similar to members of contrast categories. This is achieved by aggregating similarity across known exemplars differently according to class membership. The aggregated similarity $a$ between generation candidate $y$ and stored exemplars $x$ is given by:

@@

However, PAKCKER is only one possible exemplar-based account of category generation within our proposed framework. PACKER places specific constraints on the possible parameter values (i.e., $\phi <0$ and $\gamma > 0$). Other exemplar-based category generation models with drastically different behavior can be formalized in this framework by changing the constrains on the $\phi$ and $\gamma$ parameters. Furthermore, by fitting these parameters to a dataset, one can describe the relative roles of between-class contrast and within-class similarity in generation. For example, as will be discussed in more detail below, PACKER is formally equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic} when $\phi = 0$ and $\gamma = 1$. Likewise, when $\phi < 0$ and $\gamma = 0$, PACKER represents a contrast-only generation mode, relying exclusively on contrast when generating new categories. Finally, when $\phi = \gamma < 0$, PACKER represents a ``pure-packing'' approach, placing exemplars in any unoccupied area of the domain. Thus, beyond specifying PACKER's core proposal, the $\phi$ and $\gamma$ allow the us to address a wide variety of qualitatively distinct generation strategies.

\nbcnote{That last paragraph was added since your last read, joe!}

@@

It is worth noting that PACKER represents just one of many possible models that incorporate contrast in category generation. For example, although it may be possible to extend \cite{jern2013probabilistic}'s hierarchical sampling model to include contrast, but this mechanism would be at odds with a conjugate hierarchical Bayesian framework. In fact, we will discuss integrating our approaches in the General Discussion. In contrast, these ideas emerge naturally from the exemplar view: we have not modified any of the core elements of the GCM in defining the PACKER model, we simply aggregated similarity slightly differently. \nbcnote{That line about importance sampling can go here.} \jlanote{I think I'll write a subsection about it/include it in the subsection on future directions. I may do a bit of math to show how it would be possible too.}

@@

In this paper, we report simulations using this copy-and-tweak model. This model fits within the exemplar-based category generation framework defined above, under the constraints that $\phi = 0$ and $\gamma = 1$, and is a continuous-dimension adaptation of the model tested by \cite{jern2013probabilistic}. By formalizing a model family where PACKER and copy-and-tweak are different parameterizations of models within the same framework, the comparison between PACKER and copy-and-tweak provides a test of the explanatory value of the contrast mechanism: the account provided by copy-and-tweak will only equal that of PACKER if the contrast mechanism does not offer an advantage (i.e., if $\phi \neq 0$ significantly improves model fits).

@@

>>>>

\cite{jern2013probabilistic} recently showed that creative generation could be studied in a more controlled manner through the well-developed methods of an artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In Experiments 3 and 4 of their article, participants were exposed to members of experimenter-defined categories of "crystals" varying in size, hue, and saturation. Following a training phase during which the experimenter-defined categories were learned, participants were asked to generate novel categories of crystals. In a finding mirroring that of the \cite{ward1994structured} studies, \citeauthor{jern2013probabilistic} found that participants generated categories with the same distributional properties as the experimenter-defined categories: for example, after training on categories with a positive correlation between the size and saturation features (larger sized crystals were more saturated), participants generated novel categories with the same positive correlation. This finding is notable, as it demonstrates that category generation can be studied in a well-known and highly controlled experimental paradigm.

% % ----- This point is not true -----
% Note that generated exemplars for the novel category automatically contrasted from previous categories because the participants selected the size and saturation, but the exemplar's hue was set (and fixed) to a novel value. As such, hue was ignored in their analyses and may have satisfied any constraints contrast may have on category generation.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that broadly resembles the earlier ``Path of Least Resistance'' account. The core proposal is that participants generate new items by copying stored examples from memory and tweaking them to generate something new. The copy-and-tweak model differs from the path of least resistance account in that it notably omits the hierarchical organization of categories, as well as selectivity in which features are changed. Instead, their copy-and-tweak model corresponds to a direct exemplar-similarity approach \citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}: The model generates new items according to their similarity to known members of the target category. This model provided a poor account of the observed data, as the experiments devised by \citeauthor{jern2013probabilistic} were specifically designed to challenge this model. However, its application is notable as a first step toward explaining category generation using well-known formal approaches from the categorization literature.

@@

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims little about how generated categories should contrast with what is already known. In their simulations, the model was only tested on generation after the learner had been exposed to members of the target category, and so the model's ability to generate a new category from scratch was not evaluated. However, the model's generation is based exclusively on similarity to known members of the target category; when there are no members of the target category, generation is presumably random.

@@

Both PACKER and the GCM simulate categorization under the assumption that learners represent categories as a collection of exemplars, corresponding to the labeled stimuli they have observed. The exemplars are encoded within a $k$-dimensional psychological space, and model performance is based on the amount of similarity between the item to be categorized and the stored exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is computed as an inverse exponential function of distance \citep[following][]{attneave1950,shepard1957stimulus,shepard1987toward}:

\begin{equation}
s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k \left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, attention will be distributed uniformly in our simulations (unless otherwise noted). The value of $r$ depends on the nature of the experimental conditions being simulated: $r=1$ is appropriate for separable dimensions, whereas $r=2$ is appropriate for integral dimensions \citep[see][]{shepard1964attention,garner1974processing}. In our simulations, we set $r=1$ due to the separable nature of the stimulus dimensions used in our experiments (see Figure \ref{fig:e1-conditions}).

PACKER (as well as its name) was in part inspired by earlier work from the categorization literature \citep{hidaka2011packing}. They argued that natural categories "pack" the values of features such that different categories fill the space with distance between one another, while maintaining items within the same category close together. Inspired by this idea, PACKER proposes that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories: the most desirable generation candidates are similar to members of the target category and not similar to members of contrast categories. This is achieved by aggregating similarity across known exemplars differently according to class membership. The aggregated similarity $a$ between generation candidate $y$ and stored exemplars $x$ is given by:

@@

However, PACKER is only one possible exemplar-based account of category generation within our proposed framework. PACKER places specific constraints on the possible parameter values (i.e., $\phi <0$ and $\gamma > 0$). Other exemplar-based category generation models with drastically different behavior can be formalized in this framework by changing the constraints on the $\phi$ and $\gamma$ parameters. Furthermore, by fitting these parameters to a dataset, one can describe the relative roles of between-class contrast and within-class similarity in generation. For example, as will be discussed in more detail below, PACKER is formally equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic} when $\phi = 0$ and $\gamma = 1$. Likewise, when $\phi < 0$ and $\gamma = 0$, PACKER represents a contrast-only generation mode, relying exclusively on contrast when generating new categories. Finally, when $\phi = \gamma < 0$, PACKER represents a ``pure-packing'' approach, placing exemplars in any unoccupied area of the domain. Thus, beyond specifying PACKER's core proposal, the $\phi$ and $\gamma$ parameters allow the us to address a wide variety of qualitatively distinct generation strategies.

@@

It is worth noting that PACKER represents just one of many possible models that incorporate contrast in category generation. For example, although it may be possible to extend \cite{jern2013probabilistic}'s hierarchical sampling model to include contrast, but this mechanism would be at odds with a conjugate hierarchical Bayesian framework. In fact, we will discuss integrating our approaches in the General Discussion. In contrast, these ideas emerge naturally from the exemplar view: we have not modified any of the core elements of the GCM in defining the PACKER model, we simply aggregated similarity slightly differently.

@@

In this paper, we report simulations using this copy-and-tweak model. This model fits within the exemplar-based category generation framework defined above, under the constraints that $\phi = 0$ and $\gamma = 1$, and is a continuous-dimension adaptation of the model tested by \cite{jern2013probabilistic}. By formalizing a model family where PACKER and copy-and-tweak are different parameterizations of models within the same framework, the comparison between PACKER and copy-and-tweak provides a test of the explanatory value of the contrast mechanism: the account provided by copy-and-tweak will only equal that of PACKER if the contrast mechanism does not offer an advantage (i.e., if $\phi < 0$ significantly improves model fits).

@@

\nbcnote{Make the point that we’re explaining category generation using well known ideas, and that exemplar models are viewed as highly principled.}

\nbcnote{Discuss combining hierarchical sampling and contrast approaches.}

\nbcnote{Joe says he'll put importance sampling in future directions.}
